Refactoring \texttt{nls()}: a constructive critique

### 2023-03-02 RefactoringNLS/JNAB230203.tex
adding files needed to develop first R-Journal submisison

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Refactoring nls(): a constructive critique},
  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
  colorlinks=true,
  linkcolor={red},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

\title{Refactoring \texttt{nls()}: a constructive critique}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2022-6-15}

\begin{document}
\maketitle

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Based on work for a Google Summer of Code project ``Improvements to
\texttt{nls()}'', we consider the features and limitations of the R
function \texttt{nls()} in the context of improving and rationalizing R
tools for nonlinear regression.

Important considerations are the usability and maintainability of the
code base that provides the functionality \texttt{nls()} claims to
offer. Our work suggests that the existing code makes maintenance and
improvement very difficult, with legacy applications and examples
blocking some important updates. Discussion of these matters is relevant
to improving R generally as well as its nonlinear estimation tools.

@@

\texttt{nls()} is the tool in base R, the primary software software
distribution from the Comprehensive R Archive Network
(\emph{\url{https://cran.r-project.org}}), for estimating nonlinear
statistical models. The function dates to the 1980s and the work related
to Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

A previous issue with \texttt{nls()} that prevented it from providing
parameter estimates for zero-residual (i.e., perfect fit) data was
corrected thanks to suggestions by one of us.

@@

This issue can be easily resolved. The ``termination test'' for the
\textbf{program} rather than for ``convergence'' of the underlying
\textbf{algorithm} is the Relative Offset Convergence Criterion (see
Bates, Douglas M. and Watts, Donald G.
(\protect\hyperlink{ref-BatesWatts81}{1981})). This projects the
proposed step in the parameter vector on the gradient and estimates how
much the sum of squares loss function should decrease. This estimate is
divided by the current size of the loss function to avoid scale issues.
When we have ``converged'', the estimated decrease is very small.
However, with small residuals, the sum of squares loss function is
(almost) zero and we get the possibility of a zero-divide failure.

@@

The convergence criterion above, the principal termination control of
\texttt{nls()}, leaves out some possibilities that could be useful for
some problems. The package \texttt{nlsr} (John C Nash and Duncan Murdoch
(\protect\hyperlink{ref-nlsr2019manual}{2019})) already offers both the
safeguarded relative offset test (\textbf{roffset}) as well as a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

@@

gets over 4000 hits that are spread over some years. This could be
because the Jacobian is poorly approximated (see \textbf{Jacobian
computation} below). However, it is common in nonlinear least squares
computations that the Jacobian is very close to singular for some values
of the model parameters. In such cases, we need to find an alternative
algorithm to the Gauss-Newton iteration of \texttt{nls()}. The most
common work-around is the Levenberg-Marquardt stabilization (see
Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}), Levenberg
(\protect\hyperlink{ref-Levenberg1944}{1944}), Nash
(\protect\hyperlink{ref-jn77ima}{1977})), and versions of it have been
implemented in packages \texttt{minpack.lm} and \texttt{nlsr}. In our
work, we prepared experimental prototypes of \texttt{nls} that
incorporate stabilization, but integration with all the features of
\texttt{nls()} has proved difficult.

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{reala }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{; realb }\OtherTok{\textless{}{-}} \DecValTok{5}\NormalTok{; realc }\OtherTok{\textless{}{-}} \FloatTok{0.5}\NormalTok{; realr }\OtherTok{\textless{}{-}} \FloatTok{0.7}\NormalTok{; realm }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# underlying parameters}
\NormalTok{x }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{11} \CommentTok{\# x values; 11 data points}
\NormalTok{y }\OtherTok{\textless{}{-}}\NormalTok{ reala }\SpecialCharTok{+}\NormalTok{ realb }\SpecialCharTok{*}\NormalTok{ realr}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ realm) }\SpecialCharTok{+}\NormalTok{ realc }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# linear + exponential function}
\NormalTok{testdat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(x, y) }\CommentTok{\# save in a data frame}
\NormalTok{strt }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{, }\AttributeTok{b =} \DecValTok{5}\NormalTok{, }\AttributeTok{c =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{r =} \FloatTok{0.7}\NormalTok{, }\AttributeTok{m =} \DecValTok{1}\NormalTok{) }\CommentTok{\# give programs a good start}
\NormalTok{jform }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a }\SpecialCharTok{+}\NormalTok{ b }\SpecialCharTok{*}\NormalTok{ r}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ m) }\SpecialCharTok{+}\NormalTok{ c }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# Formula}
\FunctionTok{library}\NormalTok{(nlsr)}
\NormalTok{linexp2 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =} \ConstantTok{FALSE}\NormalTok{))}
\NormalTok{linexp2 }\CommentTok{\# Note singular values of Jacobian in rightmost column}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## residual sumsquares =  0  on  11 observations
##     after  1    Jacobian and  1 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## a                     -3            NA         NA         NA           0       26.49  
## b                      5            NA         NA         NA           0       9.615  
## c                    0.5            NA         NA         NA           0       2.466  
## r                    0.7            NA         NA         NA           0      0.1098  
## m                      1            NA         NA         NA           0   1.311e-16
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# We get an error with nls()}
\NormalTok{linexp }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =} \ConstantTok{FALSE}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nlsModel(formula, mf, start, wts, scaleOffset = scOff, nDcentral = nDcntr) : 
##   singular gradient matrix at initial parameter estimates
\end{verbatim}

@@

\begin{itemize}
\item
  A pasture regrowth problem (Huet et al.
  (\protect\hyperlink{ref-Huet2004}{2004}), page 1, based on Ratkowsky
  (\protect\hyperlink{ref-Ratkowsky1983}{1983})) has a poorly
  conditioned Jacobian and \texttt{nls()} fails with ``singular
  gradient''. Worse, numerical approximation to the Jacobian give the
  error ``singular gradient matrix at initial parameter estimates'' for
  \texttt{minpack.lm::nlsLM} so that the Marquardt stabilization is
  unable to take effect, while the analytic derivatives of
  \texttt{nlsr::nlxb} give a solution.
\item
  Karl Schilling (private communication) provided an example where a
  model specified with the formula
  \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes
  \texttt{nlsr::nlxb} to fail because the partial derivative w.r.t.
  \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for
  which \texttt{x\ =\ 0}, this is undefined. In such cases, we observed
  that \texttt{nls()} and \texttt{minpack.lm::nlsLM} found a solution,
  though it can be debated whether such lucky accidents can be taken as
  an advantage.
\end{itemize}

Note that the \texttt{selfStart} models in the base R
\texttt{./src/library/stats/R/zzModels.R} file and in package
\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
may provide the Jacobian in the ``gradient'' attribute of the
``one-sided'' formula that defines each model, and these Jacobians may
be the analytic forms. The \texttt{nls()} function, after computing the
``right hand side'' or \texttt{rhs} of the residual, then checks to see
if the ``gradient'' attribute is defined, and, if not, uses
\texttt{numericDeriv} to compute a Jacobian into that attribute. This
code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
functions. The use of analytic Jacobians almost certainly contributes to
the good performance of \texttt{nls()} on \texttt{selfStart} models.

@@

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{\texttt{selfStart}} models. There are
a number of these in base R (\texttt{./src/library/stats/R/zzModels.R})
and others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{\texttt{selfStart}} codes entangle the calculation of starting
values for parameters with the particulars of the \texttt{nls()} code.
Though there is a \texttt{getInitial()} function, this is not easy to
use to simply compute the initial parameter estimates outside of
\texttt{nls()}, in part because it may call that function. Such circular
references are, in our view, dangerous. Moreover, we believe that it
would be helpful to have \texttt{selfStart} models that allow users to
explicitly provide values other than those suggested for the starting
parameters.

@@

By providing starting values that are likely to be reasonable and by
including Jacobian code in the \texttt{selfStart} model code, some of
the deficiencies of the Gauss-Newton algorithm are mitigated. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

@@

\begin{verbatim}
              -----------------------------------------------------------
\end{verbatim}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5}\SpecialCharTok{\^{}}\NormalTok{tt }\CommentTok{\# simple weights}
\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt)}\SpecialCharTok{/}\NormalTok{scal))}
\NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# UNWEIGHTED}
\NormalTok{nowt}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Nonlinear regression model
##   model: weed ~ SSlogis(tt, Asym, xmid, scal)
##    data: parent.frame()
##   Asym   xmid   scal 
## 196.19  12.42   3.19 
##  residual sum-of-squares: 2.59
## 
## Number of iterations to convergence: 0 
## Achieved convergence tolerance: 1.12e-06
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
##  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
## attr(,"gradient")
##           Asym     xmid    scal
##  [1,] 0.027117  -1.6229  5.8103
##  [2,] 0.036737  -2.1769  7.1111
##  [3,] 0.049596  -2.8997  8.5628
##  [4,] 0.066645  -3.8266 10.1000
##  [5,] 0.089005  -4.9881 11.6015
##  [6,] 0.117921  -6.3988 12.8762
##  [7,] 0.154635  -8.0418 13.6607
##  [8,] 0.200186  -9.8498 13.6432
##  [9,] 0.255106 -11.6901 12.5267
## [10,] 0.319083 -13.3660 10.1313
## [11,] 0.390688 -14.6444  6.5083
## [12,] 0.467334 -15.3139  2.0038
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
\NormalTok{usewt}
\end{Highlighting}
\end{Shaded}

>>>>

% !TeX root = RJwrapper.tex
\title{Refactoring \texttt{nls()}: a constructive critique}
\author{by John C. Nash and Arkajyoti Bhattacharjee}

\maketitle

\abstract{%
Based on work for a Google Summer of Code project ``Improvements to \texttt{nls()}'', we consider the features and limitations of the R function \texttt{nls()} in the context of improving and rationalizing R tools for nonlinear regression. Important considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} claims to offer. Our work suggests that the existing code makes maintenance and improvement very difficult, with legacy applications and examples blocking some important updates. Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
}

@@

\texttt{nls()} is the tool in base R, the primary software software distribution
from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}),
for estimating nonlinear statistical models. The function dates to the 1980s and
the work related to Bates and Watts (1988) in S (see \emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

@@

A previous issue with \texttt{nls()} that prevented it from providing parameter
estimates for zero-residual
(i.e., perfect fit) data was corrected using a suggestion by one of us.

@@

This issue can be easily resolved. The ``termination test'' for the \textbf{program} rather than for
``convergence'' of the underlying \textbf{algorithm} is the Relative Offset
Convergence Criterion (see Bates, Douglas M. and Watts, Donald G. (1981)). This projects the proposed step
in the parameter vector on the gradient
and estimates how much the sum of squares loss function should decrease.
This estimate is divided
by the current size of the loss function to avoid scale issues.
When we have ``converged'', the estimated
decrease is very small. However, with small residuals,
the sum of squares loss function is (almost) zero and we get the possibility of a
zero-divide failure.

@@

The convergence criterion above, the principal termination control of \texttt{nls()},
leaves out some possibilities that could be useful for some problems. The
package \texttt{nlsr} (John C Nash and Duncan Murdoch (2019)) already offers both
the safeguarded relative offset test (\textbf{roffset}) as well as a \textbf{small sum of
squares} test (\textbf{smallsstest}) that compares the latest evaluated sum of squared
(weighted) residuals to \texttt{e4} times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is approximately 2.43e-55.

@@

\noindent
gets over 4000 hits that are spread over some years. This could be because the
Jacobian is poorly approximated (see \textbf{Jacobian computation} below). However,
it is common in nonlinear least squares computations that the Jacobian is very close
to singular for some values of the model parameters. In such cases, we need to find
an alternative algorithm to the Gauss-Newton iteration of \texttt{nls()}. The most common
work-around is the Levenberg-Marquardt stabilization (see Marquardt (1963),
Levenberg (1944), Nash (1977)), and versions of it have been implemented in
packages \texttt{minpack.lm} and \texttt{nlsr}.
In our work, we prepared experimental prototypes of \texttt{nls} that incorporate
stabilization, but integration with all the features of \texttt{nls()} has proved difficult.

@@

\begin{verbatim}
#> residual sumsquares =  0  on  11 observations
#>     after  1    Jacobian and  1 function evaluations
#>   name            coeff          SE       tstat      pval      gradient    JSingval   
#> a                     -3            NA         NA         NA           0       26.49  
#> b                      5            NA         NA         NA           0       9.615  
#> c                    0.5            NA         NA         NA           0       2.466  
#> r                    0.7            NA         NA         NA           0      0.1098  
#> m                      1            NA         NA         NA           0   9.798e-17
\end{verbatim}

\begin{verbatim}
#> Error in nlsModel(formula, mf, start, wts, scaleOffset = scOff, nDcentral = nDcntr) : 
#>   singular gradient matrix at initial parameter estimates
\end{verbatim}

@@

\begin{itemize}
\item
  A pasture regrowth problem (Huet et al. (2004), page 1, based on Ratkowsky (1983)) has a poorly
  conditioned Jacobian and \texttt{nls()} fails with ``singular gradient''.
  Worse, numerical approximation to the Jacobian give the error
  ``singular gradient matrix at initial parameter estimates'' for \texttt{minpack.lm::nlsLM}
  so that the Marquardt stabilization is unable to take effect, while the analytic
  derivatives of \texttt{nlsr::nlxb} give a solution.
\item
  Karl Schilling (private communication) provided an example where a model specified
  with the formula \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes \texttt{nlsr::nlxb} to fail because the partial
  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
  this is undefined. In such cases, we observed that \texttt{nls()} and \texttt{minpack.lm::nlsLM}
  found a solution, though it can be debated whether such lucky accidents can be
  taken as an advantage.
\end{itemize}

Note that the \texttt{selfStart} models in the base R \texttt{./src/library/stats/R/zzModels.R}
file and in package \texttt{nlraa} (Miguez (2021)) may
provide the Jacobian in the ``gradient'' attribute of the ``one-sided'' formula that
defines each model, and these Jacobians may be the analytic forms. The \texttt{nls()}
function, after computing the ``right hand side'' or \texttt{rhs} of the residual, then
checks to see if the ``gradient'' attribute is defined, and, if not, uses
\texttt{numericDeriv} to compute a Jacobian
into that attribute. This code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
functions. The use of analytic Jacobians almost certainly contributes to the good
performance of \texttt{nls()} on \texttt{selfStart} models.

@@

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. \texttt{nls()} offers a special class of modeling functions called
\textbf{\texttt{selfStart}} models. There are a number of these in base R
(\texttt{./src/library/stats/R/zzModels.R})
and others in R packages such as CRAN package \texttt{nlraa} (Miguez (2021)), as well as
the now-archived package \texttt{NRAIA}. Unfortunately, some \textbf{\texttt{selfStart}} codes
entangle the calculation of starting values for parameters
with the particulars of the \texttt{nls()} code. Though there is a \texttt{getInitial()} function,
this is not easy to use to simply compute the initial parameter estimates outside of
\texttt{nls()}, in part
because it may call that function. Such circular references are, in our view,
dangerous. Moreover, we believe that it would be helpful to have \texttt{selfStart}
models that allow users to explicitly provide values other than those suggested
for the starting parameters.

@@

By providing starting values that are likely to be reasonable and by including
Jacobian code in the \texttt{selfStart} model code, some of the deficiencies of the
Gauss-Newton algorithm are mitigated. However, creating such functions is a lot
of work, and their
documentation (file \texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated.
We believe that the focus should be placed on getting good initial parameters,
that is \texttt{getInitial()} function, though avoiding the current calls back to \texttt{nls()}.
Interactive tools, such as ``visual fitting'' (Nash and Velleman (1996)) might be
worth considering.

@@

@@

\begin{verbatim}
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12
weeddf <- data.frame(tt, weed)
wts <- 0.5^tt # simple weights
frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
Asym<-1; xmid<-1; scal<-1
nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
nowt
\end{verbatim}

\begin{verbatim}
#> Nonlinear regression model
#>   model: weed ~ SSlogis(tt, Asym, xmid, scal)
#>    data: parent.frame()
#>   Asym   xmid   scal 
#> 196.19  12.42   3.19 
#>  residual sum-of-squares: 2.59
#> 
#> Number of iterations to convergence: 0 
#> Achieved convergence tolerance: 1.12e-06
\end{verbatim}

\begin{verbatim}
nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
\end{verbatim}

\begin{verbatim}
#>  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
#>  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
#> attr(,"gradient")
#>           Asym     xmid    scal
#>  [1,] 0.027117  -1.6229  5.8103
#>  [2,] 0.036737  -2.1769  7.1111
#>  [3,] 0.049596  -2.8997  8.5628
#>  [4,] 0.066645  -3.8266 10.1000
#>  [5,] 0.089005  -4.9881 11.6015
#>  [6,] 0.117921  -6.3988 12.8762
#>  [7,] 0.154635  -8.0418 13.6607
#>  [8,] 0.200186  -9.8498 13.6432
#>  [9,] 0.255106 -11.6901 12.5267
#> [10,] 0.319083 -13.3660 10.1313
#> [11,] 0.390688 -14.6444  6.5083
#> [12,] 0.467334 -15.3139  2.0038
\end{verbatim}

\begin{verbatim}
usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
usewt
\end{verbatim}

\begin{verbatim}
#> Nonlinear regression model
#>   model: weed ~ SSlogis(tt, Asym, xmid, scal)
#>    data: parent.frame()
#>   Asym   xmid   scal 
#> 199.86  12.50   3.19 
#>  weighted residual sum-of-squares: 0.0196
#> 
#> Number of iterations to convergence: 2 
#> Achieved convergence tolerance: 8.29e-06
\end{verbatim}

\begin{verbatim}
usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
\end{verbatim}

\begin{verbatim}
#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
#> attr(,"gradient")
#>           Asym     xmid    scal
#>  [1,] 0.026498  -1.6157  5.8228
#>  [2,] 0.035901  -2.1679  7.1333
#>  [3,] 0.048474  -2.8890  8.6006
#>  [4,] 0.065153  -3.8149 10.1617
#>  [5,] 0.087046  -4.9775 11.6983
#>  [6,] 0.115387  -6.3933 13.0222
#>  [7,] 0.151425  -8.0483 13.8709
#>  [8,] 0.196222  -9.8787 13.9297
#>  [9,] 0.250362 -11.7553 12.8918
#> [10,] 0.313611 -13.4827 10.5608
#> [11,] 0.384641 -14.8251  6.9663
#> [12,] 0.460954 -15.5631  2.4357
\end{verbatim}

\begin{verbatim}
source("nlsModel.R")
nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
\end{verbatim}

\begin{verbatim}
#>  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
#> [11] 1.6572 1.4214
\end{verbatim}

\begin{verbatim}
nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
nmod$resid()
\end{verbatim}

\begin{verbatim}
#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
\end{verbatim}
